---
title: "capstone_movielens"
author: "Savas turkoglu"
date: "5/15/2020"
output:
  pdf_document: default
 # html_document: default
---

##1. Executive summary
This project created for Edx HarcardX data Science Certification Program capstone project.
This project tries to create a model  that estimate people rating by provided dataset that 10m movies and ratings


This document contains data analysis and prediction models.  

The next section shows the results of the previous process and then, the conclusions of the project are given.

### Code provided by the edx staff to download an create **edx** dataset.
```{r download, message = FALSE, warning = FALSE}
#Create test and validation sets
# Create edx set, validation set, and submission file
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)
ratings <- read.table(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                      col.names = c("userId", "movieId", "rating", "timestamp"))
movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                           title = as.character(title),
                                           genres = as.character(genres))
movielens <- left_join(ratings, movies, by = "movieId")
# Validation set will be 10% of MovieLens data
set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]
# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")
# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)
rm(dl, ratings, movies, test_index, temp, movielens, removed)
```
```{r aditional_libs, message = FALSE, warning = FALSE, echo = FALSE}
if(!require(anytime)) install.packages("anytime", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
```
```{r libs, message = FALSE, warning = FALSE}
library(caret)
library(anytime)
library(tidyverse)
library(lubridate)
```  

1. The Dataset
The data set contains 9000055 observations of 6 variables.  

- `userId`: Unique identification number given to each user.  -numneric
- `movieId`: Unique identification number given to each movie. -numeric
- `timestamp`: rating date and time datestams  -integrer
- `title`: Title of the movies.  -numerşc
- `genres`: Film categories. -char
- `rating`: Rating given by the user  -char 

2. Analysis Section

2 a Data preperation

Edx gave us prepared data but wee nedd some preperation 
we have to extract datestams to human readable format and we have to get year and month from date

we have to sepera title into the film name and release year

we have to split genres column 



```{r formats, message = FALSE}

# my computer sources was not enough for proccess 10m rows data,
#therefore i used  dataset partially during development

 edx <- head(edx,1000000)  #if you get memory error use this code


```

Data preparation
Edx gave us prepared data but we need some more preparation !! 

First we have to dealing with date 

Converting date-time to human readable 


```{r , message = FALSE}

edx$date <- as.POSIXct(edx$timestamp, origin="1970-01-01")
validation$date <- as.POSIXct(validation$timestamp, origin="1970-01-01")

## get year and month from date

edx$rate_year <- format(edx$date,"%Y")
edx$rate_month <- format(edx$date,"%m")

validation$rate_year <- format(validation$date,"%Y")
validation$rate_month <- format(validation$date,"%m")


```

now wee'll deal with title we have to separate title and release year  

```{r , message = FALSE}

## + in edx datasets

edx <- edx %>%
  mutate(title = str_trim(title)) %>%
  extract(title, c("title_", "release"), regex = "^(.*) \\(([0-9 \\-]*)\\)$", remove = F) %>%
  mutate(release = if_else(str_length(release) > 4,
                           as.integer(str_split(release, "-",
                                                simplify = T)[1]),
                           as.integer(release))
  ) %>%
  mutate(title = if_else(is.na(title_), title, title_)
  )
 edx <- edx %>% select(-title_)

 ## + in validaton datasets
 
 validation <- validation %>%
   mutate(title = str_trim(title)) %>%
   extract(title, c("title_", "release"), regex = "^(.*) \\(([0-9 \\-]*)\\)$", remove = F) %>%
   mutate(release = if_else( str_length(release) > 4,
                             as.integer(str_split(release, "-",
                             simplify = T)[1]),
                             as.integer(release))) %>%
   mutate(title = if_else(is.na(title_), title, title_))
 
 validation <- validation %>% select(-title_)
 
``` 


 we have to  modify the genres vars in the edx & validation dataset by column_separated 
 
```{r , message = FALSE}
 
 
 ##split daa
 
 genre_split_edx  <- edx  %>% separate_rows(genres, sep = "\\|")
 genre_split_valid <- validation  %>% separate_rows(genres, sep = "\\|")

 genres<- genre_split_edx %>%
   group_by(genres) %>%
   summarize(count = n()) 
 
 genres<- genres  %>% arrange(desc(count))
 genres
 
 genre_names <- genres$genres
 



## remove unnecessary column from datasets
 
 edx <- edx %>% select(-date, -timestamp)
 
```

```{r head}
head(edx)
```  

2.b  Data Exploratory Analysis. 



DATA OVERVIEW 



```{r rating_summary, message = FALSE}
##  take look to data by date 
edx_year <- edx %>% group_by(rate_year) %>% 
     summarize(total_rate = sum(rating), count = n()) 

 ## sort by total rate
 
 edx_year %>% arrange(desc(total_rate)) %>% knitr::kable()

 ## plot date - raiting count                     
 ggplot(edx_year, aes(y=count, x=rate_year)) + 
     geom_bar(stat = "identity")+
     labs(title = "Rate count by year ", x = "Year", y = "Raiting number")
 


## take look by rate  (1, 1.5, 2)
 edx_by_rating <- edx %>% group_by(rating) %>% 
   summarize(total_rate = sum(rating), count = n()) 
 

 
 ## plot rate - raiting count                     
 ggplot(edx_by_rating, aes(x=rating, y=total_rate)) + 
   geom_bar(stat = "identity")+
   labs(title = "Rate count by ratig ", x = "Rate", y = "Raiting number")
 
 
 
 ## rateing count by release year 
 
 movies_per_year <- edx %>%
   select(movieId, release) %>% 
   group_by(release) %>% 
   summarise(count = n())  %>%
   arrange(release)
 
 # plot
 movies_per_year %>%
   ggplot(aes(x = release, y = count)) +
   geom_line(color="blue")
 


 
 
 ## take look by genre
 edx_by_genre <- genre_split_edx %>% group_by(genres) %>% 
    summarize(total_rate = sum(rating), count = n()) 
 
 
 
 ## plot genre  - raiting count                     
 ggplot(edx_by_genre, aes(y=count, x=genres)) + 
    geom_bar(stat = "identity")+
    labs(title = "Rate count by Genre ", x = "Genre", y = "Raiting number")
 
 
```




3-Modelling



```{r create_train_test, message = FALSE, warning = FALSE}
    
# we'll use this RMSE 
 RMSE <- function(true_ratings, predicted_ratings){
    sqrt(mean((true_ratings - predicted_ratings)^2,na.rm = T))
 }
 
 
 ## Dataset’s mean rating is used to predict the same rating for all movies
 mu <- mean(edx$rating)  
 mu
 
 
 ### BASELINE MODEL ###
 baseline_model_rmse <- RMSE(validation$rating,mu)
 ## Test results 
 baseline_model_rmse
 
 ## create table that we vcollect methods and RMSEs 
 
 rmse_results <- data_frame(method = "Baseline model", RMSE =  baseline_model_rmse)
 
 #check
 rmse_results
 
 
 
 ### MOVIE  MODEL ###
 #rating  are different effected from different movies
 
 
 movie_av <- edx %>% 
    group_by(movieId) %>% 
    summarize(b_i = mean(rating - mu))
 
 
 #rediction
 
 pred_movie_av <- validation %>% 
    left_join(movie_av, by='movieId') %>%
    mutate(pred = mu + b_i) 
 
 
 # calculate RMSE
 model_movie_av_rmse <- RMSE(validation$rating,pred_movie_av$pred)
 model_movie_av_rmse
 
 # add to table
 rmse_results <- bind_rows(rmse_results,
                           data_frame(method="Movie Effect Model",  
                                      RMSE =  model_movie_av_rmse ))
 # check results
 rmse_results %>% knitr::kable()
```
 
 
 
 USER AND MOVIE MODEL 
 
 different users can interest different movies and this could effect to rating
 
 
```{r create_train_test_, message = FALSE, warning = FALSE}
 
 user_av <- edx %>% 
    left_join(movie_av, by='movieId') %>%
    group_by(userId) %>%
    summarize(b_u = mean(rating - mu - b_i))
 

 # prediction
 pred_movie_user <- validation %>% 
    left_join(movie_av, by='movieId') %>%
    left_join(user_av, by='userId') %>%
    mutate(pred = mu + b_i + b_u) %>% .$pred
 
 
 # calculate RMSE
 model_user_movie_rmse <- RMSE(validation$rating,  pred_movie_user)
 
 rmse_results <- bind_rows(rmse_results,
                           data_frame(method="User & Movie Model",  
                                      RMSE = model_user_movie_rmse ))
 rmse_results %>% knitr::kable()

```
 
 ###  REGULARİZATİON USİNG USER AND MOVİE  ####
 
 # We have learned on Edx  Machine learning course 
 #Regularization is very usefull method  to imprıve our results we'll use this method 
 #using user and movie
 
```{r create_train_test_reg, message = FALSE, warning = FALSE}
 
 
 lambdas <- seq(0, 10, 0.25)
 
 # this loop will take time
 rmses <- sapply(lambdas, function(l){
    
    mu <- mean(edx$rating)
    
    b_i <- edx %>% 
       group_by(movieId) %>%
       summarize(b_i = sum(rating - mu)/(n()+l))
    
    b_u <- edx %>% 
       left_join(b_i, by="movieId") %>%
       group_by(userId) %>%
       summarize(b_u = sum(rating - b_i - mu)/(n()+l))
    
    pred_rating <- validation %>% 
       left_join(b_i, by = "movieId") %>%
       left_join(b_u, by = "userId") %>%
       mutate(pred = mu + b_i + b_u) %>%
       .$pred
    
    return(RMSE(validation$rating,pred_rating))
 })
 
 # now we'ill get lowest value from rmses
 lambda <- lambdas[which.min(rmses)]
 lambda
 
 
 # now we can compute regularized estimates of b_i using this lowest lambda
 mov_av_reg <- edx %>% 
    group_by(movieId) %>% 
    summarize(b_i = sum(rating - mu)/(n()+lambda), n_i = n())
 
 # also we can compute regularized estimates of b_u using  this lowest lambda
 user_av_reg <- edx %>% 
    left_join(mov_av_reg, by='movieId') %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - mu - b_i)/(n()+lambda), n_u = n())
 
 
 # Prediction
 pred_ratings_reg <- validation %>% 
    left_join(mov_av_reg, by='movieId') %>%
    left_join(user_av_reg, by='userId') %>%
    mutate(pred = mu + b_i + b_u) %>%     .$pred
 
 
 #  get rmse
 reg_rmse <- RMSE(validation$rating,pred_ratings_reg)
 
 # add to table
 rmse_results <- bind_rows(rmse_results, 
                           data_frame(method="Regulariz User & Movie Model", 
                                      RMSE = reg_rmse ))
 
 #check results
 rmse_results %>% knitr::kable()
 
```
 
 REGULARİZATİON USER , MOVIE , YEAR, GENRE ###
 
 
```{r create_train_test_pr, message = FALSE, warning = FALSE}
 
 lambdas <- seq(0, 10, 0.5)
 # this part will take time
 rmses <- sapply(lambdas, function(l){
    
    mu <- mean(edx$rating)
    
    b_i <-  genre_split_edx %>% 
       group_by(movieId) %>%
       summarize(b_i = sum(rating - mu)/(n()+l))
    
    b_u <-  genre_split_edx %>% 
       left_join(b_i, by="movieId") %>%
       group_by(userId) %>%
       summarize(b_u = sum(rating - b_i - mu)/(n()+l))
    
    b_y <-  genre_split_edx %>%
       left_join(b_i, by='movieId') %>%
       left_join(b_u, by='userId') %>%
       group_by(release) %>%
       summarize(b_y = sum(rating - mu - b_i - b_u)/(n()+lambda), n_y = n())
    
    b_g <-  genre_split_edx %>%
       left_join(b_i, by='movieId') %>%
       left_join(b_u, by='userId') %>%
       left_join(b_y, by = 'release') %>%
       group_by(genres) %>%
       summarize(b_g = sum(rating - mu - b_i - b_u - b_y)/(n()+lambda), n_g = n())
    
    # prediction
    reg_predict <- genre_split_valid %>% 
       left_join(b_i, by='movieId') %>%
       left_join(b_u, by='userId') %>% 
       left_join(b_y, by = 'release') %>%
       left_join(b_g, by = 'genres') %>%
       mutate(pred = mu + b_i + b_u + b_y + b_g) %>% 
       .$pred
    
    return(RMSE(genre_split_valid$rating,reg_predict))
 })

 
 # now we'ill get lowest value from rmses
 lambda_all <- lambdas[which.min(rmses)]
 lambda_all
 
 
 # now we can copmute rmse with tihs lowet lambda value
 
 
 movie_reg_av_ <- genre_split_edx %>% 
    group_by(movieId) %>% 
    summarize(b_i = sum(rating - mu)/(n()+lambda_all), n_i = n())
 
 user_reg_av_ <- genre_split_edx %>% 
    left_join(movie_reg_av_, by='movieId') %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - mu - b_i)/(n()+lambda_all), n_u = n())
 
 year_reg_av_ <- genre_split_edx %>%
    left_join(movie_reg_av_, by='movieId') %>%
    left_join(user_reg_av_, by='userId') %>%
    group_by(release) %>%
    summarize(b_y = sum(rating - mu - b_i - b_u)/(n()+lambda_all), n_y = n())
 
 genre_reg_av_ <- genre_split_edx %>%
    left_join(movie_reg_av_, by='movieId') %>%
    left_join(user_reg_av_, by='userId') %>%
    left_join(year_reg_av_, by = 'release') %>%
    group_by(genres) %>%
    summarize(b_g = sum(rating - mu - b_i - b_u - b_y)/(n()+lambda_all), n_g = n())
 
 
 #prediction
 req_pred <- genre_split_valid %>% 
    left_join(movie_reg_av_, by='movieId') %>% 
    left_join(user_reg_av_, by='userId') %>%
    left_join(year_reg_av_, by = 'release') %>%
    left_join(genre_reg_av_, by = 'genres') %>%
    mutate(pred = mu + b_i + b_u + b_y + b_g) %>% 
    .$pred
 
 #compute rmse
 model_4_rmse <- RMSE(genre_split_valid$rating,req_pred)
 
 
 rmse_results <- bind_rows(rmse_results,
                           data_frame(method="Regulariztion User, Year,  Movie, Genre  Model",  
                                      RMSE = model_4_rmse ))
```



4. Results
```{r table_complete, message = FALSE, warning = FALSE}
 
    rmse_results %>% knitr::kable()

```  



##5. Conclusion
The variables `userId` and `movieId` have sufficient predictive power to permit us to predict how a user will rate a movie. Therefore, the user could decide to spend more time using the service.


My project Github repository  is **[in this link](https://github.com/savasturkoglu1/edx-project)**